---
title: "Processing label-free MS/MS data"
author: "André F. B. Menezes"
date: "`r format(Sys.time(), '%B %d, %Y')`"
toc: true
number-sections: true
bibliography: references.bib
format:
  html:
    html-math-method: katex
    code-tools: true
    self-contained: true
    code-fold: true
    code-summary: "Show the code"
execute:
  warning: false
---

```{r packages, include=FALSE}
suppressMessages(library(QFeatures))
library(ggplot2)
library(cowplot)
theme_set(
  theme_cowplot() +
    background_grid() +
    theme(legend.position = "top")
)
```


# Introduction {#sec-intro}

This report describes the processing steps use to organized the 
label-free proteomics of _Escherichia_ _coli_ cultures of bacteria data.
This data was first analyzed by @Margalit2022.

The report is organized as follows.

# Data import and management {#sec-data}

The data from @Margalit2022 experiment was first pre-processed using
[MaxQuant](https://www.maxquant.org/) software. MaxQuant provides the 
evidence.txt file which combines all the informationa about identified peptides
and is the only file required for processing the data and performing downstream
analysis.

The first step we read the evidence file, filters out contaminants and reverse
sequences, remove the Tetracycline antibiotic and pivot the matrix data in
wide format.

```{r data-import}
# Import the raw data (evidence)
data_raw <- read.delim("../data/evidence.txt")

cols <- list(
  charge = "Charge",
  sequence = 'Sequence',
  modified_sequence = 'Modified.sequence',
  modifications = 'Modifications',
  protein_group = 'Proteins',
  protein = 'Leading.Razor.Protein',
  experiment = 'Experiment',
  reverse = 'Reverse',
  contaminant = 'Contaminant',
  intensity = 'Intensity'
)
data_raw <- data_raw[, as.character(cols)]
colnames(data_raw) <- names(cols)
dplyr::glimpse(data_raw)

# Filters out contaminants and reverse sequences
data_raw <- data_raw[
  which(data_raw$contaminant != "+" & data_raw$reverse != "+"), ]

# Removing Tetracycline group
data_raw <- data_raw[!(data_raw$experiment %in% c("Tet2", "Tet3", "Tet4")), ]

# Creating a unique id, since there are multiple peptides corresponding to the
# same sequence
chosen_cols <- c("modified_sequence", "protein", "experiment", "intensity")
data_raw <- data_raw[, chosen_cols] |> 
  dplyr::group_by(experiment, modified_sequence, protein) |> 
  dplyr::mutate(number_peptides = dplyr::n()) |> 
  dplyr::ungroup() |>
  dplyr::arrange(experiment, modified_sequence, protein)
data_raw$unique_id <- seq.int(1, nrow(data_raw))

# Pivot to create the data matrix at PSMs level
pivotted_psms <- tidyr::pivot_wider(
  data = data_raw,
  id_cols = c(unique_id, number_peptides, modified_sequence, protein),
  names_from = experiment,
  values_from = intensity)
rm(data_raw)
dplyr::glimpse(pivotted_psms)
```

A useful infrastructure for omics data management in R is the
`SummarizedExperiment` class [@Morgan2022]. For quantitative proteomics data
we have the `QFeatures` package [@Gatto2022] which provides infrastructure
to process and manage the data.

In the sequel an object of `SummarizedExperiment` is created with the
intensity at PSMs level, then it is store in at `QFeatures` object. Finally,
we aggregate from PSMs to peptides level by computing the median across the
PSMs observations for each sample.

```{r qfeature-obj}
# Columns data referring the design of experiment
samples_names <- colnames(pivotted_psms)[-c(1:4)]
map_names <- list(
  "amp" = "Ampicillin",
  "cef" = "Cefotaxime",
  "imp" = "Impipenem",
  "cip" = "Ciprofloxacin",
  "cont" = "Control")
group_names <- as.character(map_names[
  tolower(gsub("[0-9]+", "", samples_names))])
col_data <- DataFrame(group = group_names,
                      replicate = rep(1L:3L, length(map_names)),
                      sample_names = samples_names,
                      row.names = paste0("Sample_",
                                         seq_len(length(group_names))))
# Creating a SummarizedExperiment for PSMs level
row_data <- pivotted_psms[, c(1:4)]
m_psms <- as.matrix(pivotted_psms[, -c(1:4)])
colnames(m_psms) <- rownames(col_data)
se_psms <- SummarizedExperiment(assays = list(intensity = m_psms),
                                rowData = row_data, colData = col_data)

# Creating the QFeature object --------------------------------------------
colnames(pivotted_psms)[5:19] <- rownames(col_data)
fts <- readQFeatures(table = pivotted_psms, ecol = 5:19, name = "psms")
assayNames(fts[["psms"]])[1L] <- "intensity"
colData(fts) <- col_data

# Aggregate data at peptide level -----------------------------------------
# rows: peptides sequences and columns: samples
fts <- aggregateFeatures(object = fts, i = "psms", fcol = "modified_sequence",
                         name = "peptides", fun = colMedians, na.rm = TRUE)
fts
colData(fts)
head(assay(fts[["peptides"]]))
rowData(fts[["peptides"]])
assayNames(fts[["peptides"]])[1L] <- "intensity"
```

In this stage, it is important to note that at PSMs level we have a intensity
matrix with dimensions `r dim(fts[["psms"]])`, while at peptides level we
have `r dim(fts[["peptides"]])`.

A common feature of label-free proteomics data is the great quantity of 
missing values due to absence of detection of a feature. For such cases,
the missing values are expected to be randomly distributed in the data, thus
are missing at random (MAR).

Note that the percentage of missing values at PSMs and peptides levels are,
respectively,
`r paste0(round(100 * mean(is.na(assay(fts[["psms"]]))), 2), "%")` and 
`r paste0(round(100 * mean(is.na(assay(fts[["peptides"]]))), 2), "%")`.

Let's inspect the missing values distribution at peptides level.

```{r vis-miss-peptides}
#| fig-height: 8
#| label: fig-missing_values_peptides
#| fig-cap: Distribution of missing values at peptides level.
data_ <- as.data.frame(assay(fts[["peptides"]]))
colnames(data_) <- fts$sample_names
naniar::vis_miss(data_)
```

Now, we will aggregate at proteins level by computing the median and examine
the distribution of missing values. The percentage of missing values at
proteins level is
`r paste0(round(100 * mean(is.na(assay(fts[["proteins"]]))), 2), "%")`.

```{r vis-miss-proteins}
#| fig-height: 8
#| label: fig-missing_values_proteins
#| fig-cap: Distribution of missing values at proteins level.
fts <- aggregateFeatures(object = fts, i = "peptides", fcol = "protein",
                         name = "proteins", fun = colMedians, na.rm = TRUE)
assayNames(fts[["proteins"]])[1L] <- "intensity"
colData(fts[["proteins"]]) <- col_data
data_ <- as.data.frame(assay(fts[["proteins"]]))
colnames(data_) <- fts$sample_names
naniar::vis_miss(data_)
```

From @fig-missing_values_peptides and @fig-missing_values_proteins it is
observed that the missing values are greater at peptides level than proteins,
as expected. This fact will guide our decision of imputation technique.

# Processing steps {#sec-processing}

Processing the raw data set is a crucial task when analyzing omics data.
Proteomics quantification data have technical variability due to experimental
process, remove those variability and keep only the biological ones is
fundamental for the downstream statistical analysis.

It is important to mention that the processing steps performs at proteins
level, since as shown in @fig-missing_values_peptides and
@fig-missing_values_proteins there are many missing values at peptides
level, which can affect proper corrections.
Furthermore, the imputation and normalization are performed at log2
transformed intensity.

## Filtering

First, note from @fig-ecdf_missing_values_proteins  that 50% of the proteins
(total of 778) do not have any missing values in all samples.
However, 21% of the proteins have more than 50% of missing values across 
all samples (see the red dot). Therefore, we will remove the proteins that
have more than 50% of missing values.

```{r filt}
#| fig-width: 8
#| label: fig-ecdf_missing_values_proteins
#| fig-cap: Cumulative distribution of the percentage of missing values at proteins level across samples.
na_proteins <- rowMeans(is.na(assay(fts[["proteins"]])))
th <- 0.50
ggplot(data = data.frame(x = na_proteins), aes(x = x)) +
  stat_ecdf() +
  geom_point(data = data.frame(x = th, y = mean(na_proteins <= th)),
             aes(x = x, y = y), col = "red", size = 4) +
  geom_rug() +
  labs(x = "Percentage of missing values", y = "Cumulative probability")

fts[["proteins"]] <- fts[["proteins"]][na_proteins < th, ]
```

## Imputation

The imputation of missing at random (MAR) values is very discussed in theory
and real applications. In the context of label-free quantitative proteomics
data @Lazar2016 performed a series of imputation methods comparison using
simulated data set. The authors showed that overall the MAR-devoted methods,
specifically _k_NN, SVD and MLE provided good performance for the imputation
task.

In this study, we consider the _k_NN method implemented in the Bioconductor
R package `impute` [@Hastie2022]. For each protein abundance, $x_i$ the method
the _k_NN method works as follows:

  1. Calculate the Euclidean distance between $\mathbf{x}_i^c$ and all other
  proteins in $\mathbf{X}^c$;
  2. Identify the $k$ closest proteins;
  3. Impute the missing values of $x_i$ using the average of the corresponding
  $k$ closest.

Here, $\mathbf{x}_i^c$ and $\mathbf{X}_i^c$ denote the complete observed values
of $x_i$ and all other proteins.


```{r imputation}
assay(fts[["proteins"]], "log2_intensity") <- log2(assay(fts[["proteins"]]))
logx_imputed <- impute::impute.knn(
  data = assay(fts[["proteins"]], "log2_intensity"))
assay(fts[["proteins"]], "log2_imputed") <- logx_imputed$data
```

## Normalization

Normalization aims to remove systematic differences due to experimental 
instrumentation, ensuring that any observed heterogeneity or differential
expression between samples are driven by biology not technical biases.
For label-free proteomics quantification data @Valikangas2018 systematic
reviewed normalization methods using spike-in data sets. The authors shows that
the faster cyclic loess normalization introduced by @Ballman2004 gives well
results for proteomics data.
We use the faster cyclic loess for normalization on the imputed log2 intensity.
This method is implemented in `normalizeCyclicLoess` function with argument 
`method = "fast"` from `limma` package [@Ritchie2015].


```{r normalization}
assay(fts[["proteins"]], "log2_normalized") <- limma::normalizeCyclicLoess(
  x = assay(fts[["proteins"]], "log2_imputed"), method = "fast")
```


# Comparison {#sec-comparison}

This section shows graphical comparison between the data set processed by
@Margalit2022 and our proposal describes below.

```{r import-margalit, include=FALSE}
se_margalit <- readRDS("../data/se_processed.rds")
rownames(se_margalit) <- rowData(se_margalit)$protein__id

# Organizing the data
pivotting_data <- function(se, i = "log2_imputed") {
  tb <- assay(se, i) |>
    t() |>
    dplyr::as_tibble() |>
    dplyr::mutate(replicate = se$replicate,
                  group = se$group,
                  variable = i) |>
    tidyr::pivot_longer(cols = -c(replicate, group, variable),
                        names_to = "proteins",
                        values_to = "value") |>
    dplyr::select(proteins, group, replicate, variable, value)
  tb
}

data_pivotted <- dplyr::bind_rows(
  pivotting_data(se = fts[["proteins"]], i = "intensity"),
  pivotting_data(se = fts[["proteins"]], i = "log2_imputed"),
  pivotting_data(se = fts[["proteins"]], i = "log2_normalized"),
  pivotting_data(se = se_margalit, i = "log_intensity")) |> 
  dplyr::mutate(variable = dplyr::case_when(
    variable == "log_intensity" ~ "Margalit et. al (2022)",
    variable == "intensity" ~ "Unprocessed intensity",
    variable == "log2_imputed" ~ "Log2 intensity imputed",
    variable == "log2_normalized" ~ "Log2 intensity normalized"),
    variable = forcats::fct_relevel(
      factor(variable), "Margalit et. al (2022)", "Unprocessed intensity",
      "Log2 intensity imputed"))
```

@fig-densities shows the proteins abundance densities colored by group and
faceted according to the transformation performed. The Unprocessed intensity
panel shows a very highly skewed distributions. The others panel present the
transformed intensity and we can see that the distribution is symmetric.
When comparing the two below panels we can observe the effect of normalization.


```{r all-densities}
#| fig-width: 8
#| label: fig-densities
#| fig-cap: Proteins abundance densities by transformation and group.
ggplot(data_pivotted, aes(x = value, fill = group, col = group)) +
  facet_wrap(~variable, scales = "free") +
  geom_density(alpha = 0.3) +
  geom_rug(show.legend = FALSE) +
  labs(x = "Abundance", y = "Density", fill = "", col = "")
```

@fig-densities_comparison gives a comparison between the densities by the
processing methods. We clearly see that the distribution by @Margalit2022
are spreader than ours, which are more symmetric.

```{r densities-comparison}
#| fig-width: 8
#| label: fig-densities_comparison
#| fig-cap: Comparison of proteins abundance densities between the processing approaches.
data_pivotted |> 
  dplyr::filter(variable %in% c("Margalit et. al (2022)",
                         "Log2 intensity normalized"),
                proteins %in% rownames(se_margalit)) |> 
  ggplot(aes(x = value, fill = variable, col = variable)) +
  facet_wrap(~group) +
  geom_density(alpha = 0.4) +
  geom_rug(show.legend = FALSE) +
  labs(x = "Abundance", y = "Density", fill = "", col = "") +
  scale_x_continuous(breaks = scales::pretty_breaks(6)) +
  scale_y_continuous(breaks = scales::pretty_breaks(6))
```

Other graphical inspection between the data process steps is provided by the
PCA plots in @fig-pca. From the second panel we can see that
the samples from the same group are closer, while the data process used by
@Margalit2022 only grouped the samples from Cipofloxaxin.

```{r pca-comparison}
#| fig-height: 8
#| label: fig-pca
#| fig-cap: PCA performed on imputed $\log_2$ intensity before and after normalization.
pca_non_normalized <- scater::calculatePCA(
  x = fts[["proteins"]], exprs_values = "log2_imputed")
pca_normalized <- scater::calculatePCA(
  x = fts[["proteins"]], exprs_values = "log2_normalized")
pca_margalit <- scater::calculatePCA(
  x = se_margalit, exprs_values = "log_intensity")

tb <- dplyr::as_tibble(pca_non_normalized) |>
  dplyr::mutate(sample = rownames(pca_non_normalized),
                group = colData(fts[["proteins"]])$group,
                measure = "Log2 intensity imputed") |>
  dplyr::bind_rows(
    dplyr::as_tibble(pca_normalized) |>
      dplyr::mutate(group = colData(fts[["proteins"]])$group,
                    measure = "Log2 intensity normalized")) |>
  dplyr::bind_rows(
    dplyr::as_tibble(pca_margalit) |>
      dplyr::mutate(group = colData(se_margalit)$group,
                    measure = "Margalit et. al (2022)"))

ggplot(tb, aes(x = PC1, y = PC2, col = group)) +
  facet_wrap(~measure, ncol = 1) +
  geom_point(size = 4) +
  labs(x = "PCA 1", y = "PCA 2", col = "") +
  scale_x_continuous(breaks = scales::pretty_breaks(8)) +
  scale_y_continuous(breaks = scales::pretty_breaks(8))
```

