---
title: "Processing label-free MS/MS data"
author: "André F. B. Menezes, Beatrice Tropea, and Cormac Monaghan"
date: "`r format(Sys.time(), '%B %d, %Y')`"
toc: true
number-sections: true
bibliography: references.bib
format:
  html:
    html-math-method: katex
    code-tools: true
    self-contained: true
    code-fold: true
    code-summary: "Show the code"
execute:
  warning: false
---

```{r packages, include=FALSE}
suppressMessages(library(QFeatures))
library(ggplot2)
library(cowplot)
theme_set(
  theme_cowplot() +
    background_grid() +
    theme(legend.position = "top")
)
```


# Introduction {#sec-intro}

This report describes the processing steps used to organize the 
label-free proteomics of _Escherichia_ _coli_ cultures bacterial data.
This data was first analyzed by @Margalit2022.

The report is organized as followed. @sec-data describes data import and management
using the QFeatures package. Processing steps, namely filtering, imputation, and
normalization are discussed in @sec-processing.
Finally, @sec-comparison presents a comparison between the processed data from
the original paper and from the steps performed below.

# Data import and management {#sec-data}

The data from @Margalit2022 experiment was first pre-processed using the
[MaxQuant](https://www.maxquant.org/) software. MaxQuant provides an 
`evidence.txt` file that combines all the information about identified peptides
and is the only file required for processing the data and performing downstream
analysis.

For the first step, we read the evidence file, filter out contaminants and reverse
sequences, remove the Tetracycline antibiotic, and pivot the matrix data in
wide format.

```{r data-import}
# Import the raw data (evidence)
data_raw <- read.delim("../data/evidence.txt")

cols <- list(
  charge = "Charge",
  sequence = 'Sequence',
  modified_sequence = 'Modified.sequence',
  modifications = 'Modifications',
  protein_group = 'Proteins',
  protein = 'Leading.Razor.Protein',
  experiment = 'Experiment',
  reverse = 'Reverse',
  contaminant = 'Contaminant',
  intensity = 'Intensity'
)
data_raw <- data_raw[, as.character(cols)]
colnames(data_raw) <- names(cols)
dplyr::glimpse(data_raw)

# Filters out contaminants and reverse sequences
data_raw <- data_raw[
  which(data_raw$contaminant != "+" & data_raw$reverse != "+"), ]

# Removing Tetracycline group
data_raw <- data_raw[!(data_raw$experiment %in% c("Tet2", "Tet3", "Tet4")), ]

# Creating a unique id, since there are multiple peptides corresponding to the
# same sequence
chosen_cols <- c("modified_sequence", "protein", "experiment", "intensity")
data_raw <- data_raw[, chosen_cols] |> 
  dplyr::group_by(experiment, modified_sequence, protein) |> 
  dplyr::mutate(number_peptides = dplyr::n()) |> 
  dplyr::ungroup() |>
  dplyr::arrange(experiment, modified_sequence, protein)
data_raw$unique_id <- seq.int(1, nrow(data_raw))

# Pivot to create the data matrix at PSM's level
pivotted_psms <- tidyr::pivot_wider(
  data = data_raw,
  id_cols = c(unique_id, number_peptides, modified_sequence, protein),
  names_from = experiment,
  values_from = intensity)
rm(data_raw)
dplyr::glimpse(pivotted_psms)
```

A useful infrastructure for omics data management in R is the
`SummarizedExperiment` class [@Morgan2022]. For quantitative proteomics data,
we have the `QFeatures` package [@Gatto2022] which provides an infrastructure
to process and manage the data.

For the second step, an object of `SummarizedExperiment` is created with the
intensity at the PSM's level. It is then stored in a `QFeatures` object. Finally,
we aggregate from PSM's to peptide level by computing the median across the
PSM's observations for each sample.

```{r qfeature-obj}
# Columns data referring the design of experiment
samples_names <- colnames(pivotted_psms)[-c(1:4)]
map_names <- list(
  "amp" = "Ampicillin",
  "cef" = "Cefotaxime",
  "imp" = "Impipenem",
  "cip" = "Ciprofloxacin",
  "cont" = "Control")
group_names <- as.character(map_names[
  tolower(gsub("[0-9]+", "", samples_names))])
col_data <- DataFrame(group = group_names,
                      replicate = rep(1L:3L, length(map_names)),
                      sample_names = samples_names,
                      row.names = paste0("Sample_",
                                         seq_len(length(group_names))))
# Creating a SummarizedExperiment for PSM#s level
row_data <- pivotted_psms[, c(1:4)]
m_psms <- as.matrix(pivotted_psms[, -c(1:4)])
colnames(m_psms) <- rownames(col_data)
se_psms <- SummarizedExperiment(assays = list(intensity = m_psms),
                                rowData = row_data, colData = col_data)

# Creating the QFeature object --------------------------------------------
colnames(pivotted_psms)[5:19] <- rownames(col_data)
fts <- readQFeatures(table = pivotted_psms, ecol = 5:19, name = "psms")
assayNames(fts[["psms"]])[1L] <- "intensity"
colData(fts) <- col_data

# Aggregate data at peptide level -----------------------------------------
# rows: peptide sequences & columns: samples
fts <- aggregateFeatures(object = fts, i = "psms", fcol = "modified_sequence",
                         name = "peptides", fun = colMedians, na.rm = TRUE)
fts
colData(fts)
head(assay(fts[["peptides"]]))
rowData(fts[["peptides"]])
assayNames(fts[["peptides"]])[1L] <- "intensity"
```

In this stage, it is important to note that at the PSM's level we have an intensity
matrix with dimensions `r dim(fts[["psms"]])`, while at the peptide level we
have dimensions `r dim(fts[["peptides"]])`.

A common feature of label-free proteomics data is the large number of 
missing values due to the absence of feature detection. For such cases,
the missing values are expected to be randomly distributed in the data, thus 
referred to as missing at random (MAR).

Note that the percentage of missing values at PSM's and peptide levels are,
respectively,
`r paste0(round(100 * mean(is.na(assay(fts[["psms"]]))), 2), "%")` and 
`r paste0(round(100 * mean(is.na(assay(fts[["peptides"]]))), 2), "%")`.
@fig-missing_values_peptides
shows us distribution of the missing values at peptide level.

```{r vis-miss-peptides}
#| fig-height: 8
#| label: fig-missing_values_peptides
#| fig-cap: Distribution of missing values at peptide level.
data_ <- as.data.frame(assay(fts[["peptides"]]))
colnames(data_) <- fts$sample_names
naniar::vis_miss(data_)
```

Now, we will aggregate at protein level by computing the median and examining
the distribution of missing values. The percentage of missing values at
protein level is
`r paste0(round(100 * mean(is.na(assay(fts[["proteins"]]))), 2), "%")`.

```{r vis-miss-proteins}
#| fig-height: 8
#| label: fig-missing_values_proteins
#| fig-cap: Distribution of missing values at protein level.
fts <- aggregateFeatures(object = fts, i = "peptides", fcol = "protein",
                         name = "proteins", fun = colMedians, na.rm = TRUE)
assayNames(fts[["proteins"]])[1L] <- "intensity"
colData(fts[["proteins"]]) <- col_data
data_ <- as.data.frame(assay(fts[["proteins"]]))
colnames(data_) <- fts$sample_names
naniar::vis_miss(data_)
```

From @fig-missing_values_peptides and @fig-missing_values_proteins we can see 
that the missing values at peptide level are greater than that of protein level
(as expected). This fact will guide our decision on the imputation technique.

# Processing steps {#sec-processing}

Processing the raw data set is a crucial task when analyzing omics data.
Proteomics quantification data have technical variability due to experimental
processes. Removing that variability and keeping only the biological variability 
is fundamental for later downstream statistical analysis.

It is important to mention that these processing steps are performed at the
protein level, since as shown in @fig-missing_values_peptides and
@fig-missing_values_proteins, there are many missing values at the peptide
level, which can affect proper corrections.
Furthermore, the imputation and normalization are performed at $\log_2$
transformed intensity.

@tbl-steps_comparison summarizes the processing steps used in @Margalit2022
and the alternative approach adopted in this project.
The imputation was performed by sampling from a Normal distribution with
downshift of 1.8 times the mean standard deviation (SD) of all measured values
and a width of 0.3 times this SD.

```{r tbl-comparison}
#| label: tbl-steps_comparison
#| tbl-cap: Comparison between the processed steps.

tab_steps_comparison <- dplyr::tibble(
  "Steps" = c("Aggregation", "Filter", "Imputation", "Normalization"),
  "Margalit et al. (2022)" = c("?", "?", "Sampling from Normal distribution",
                               "Z-score"),
  "Alternative" = c("Median", "Pct Missing $> 50%$ and Number peptides $<= 3$",
                    "$k$-NN", "Faster cyclic loess"))

kableExtra::kbl(x = tab_steps_comparison) |> 
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)

```


## Filtering

We start the processing step by filters out uninteresting proteins. We adopted
two filter. First, proteins that have less or equal 3 peptides matched.
Second, proteins with more than 50% of missing values across all samples.

@fig-number_peptides shows the distribution of the number of peptides
used to aggregate at the protein level. The figure shows that most of the
proteins are matched to 3, 4, 5 or 6 peptides.
Hence, proteins with less or equal than 3 matches peptides were exclude from
the analysis. This means that 25\% out of `r nrow(fts[["proteins"]])`
were excluded (see the red dot in @fig-number_peptides).


```{r r-vis-number-peptides}
#| fig-width: 8
#| label: fig-number_peptides
#| fig-cap: Distribution of the number of peptide per proteins.

ggplot(data = data.frame(x = rowData(fts[["proteins"]])$.n), aes(x = x)) +
  stat_ecdf() +
  geom_point(data = data.frame(x = 3, y = mean(rowData(fts[["proteins"]])$.n <= 3)),
             aes(x = x, y = y), col = "red", size = 4) +
  geom_rug() +
  labs(x = "Number of peptides per protein", y = "Cumulative probability") +
  scale_x_continuous(breaks = scales::pretty_breaks(6))
```

Second, we note that from @fig-ecdf_missing_values_proteins that 50% of the
proteins (total of 778) do not have any missing values in all samples.
However, 21% of the proteins have more than 50% missing values across 
all samples (see the red dot). Therefore, we will remove these above mentioned
proteins.

```{r filt}
#| fig-width: 8
#| label: fig-ecdf_missing_values_proteins
#| fig-cap: Cumulative distribution of the percentage of missing values at proteins level across samples.
na_proteins <- rowMeans(is.na(assay(fts[["proteins"]])))
th <- 0.50
ggplot(data = data.frame(x = na_proteins), aes(x = x)) +
  stat_ecdf() +
  geom_point(data = data.frame(x = th, y = mean(na_proteins <= th)),
             aes(x = x, y = y), col = "red", size = 4) +
  geom_rug() +
  labs(x = "Percentage of missing values", y = "Cumulative probability")

# Filtering
fts[["proteins"]] <- fts[["proteins"]][na_proteins < th, ]
fts[["proteins"]] <- fts[["proteins"]][rowData(fts[["proteins"]])$.n > 3, ]
```

## Imputation

The imputation of missing at random (MAR) values is widely discussed in both
theory and real practice. In the context of label-free quantitative proteomics
data @Lazar2016 performed a series of imputation method comparisons using
simulated and read data set. The authors showed that overall the MAR-devoted
methods, specifically _k_-NN, SVD, and MLE provided good performance for the
imputation task.

As mentioned the imputation technique used by Perseus is sampling from a 
Normal distribution with shifts in the mean and variance that mimic low
undetected intensities, therefore the missing values are filled by low values.
This imputed method could lead to an inflation in the variance producing
differences that would not appear. 


In this study, we consider the _k_-NN method implemented in the Bioconductor
R package `impute` [@Hastie2022]. For each protein abundance, $x_i$ the method
the _k_-NN method works as follows:

  1. Calculate the Euclidean distance between $\mathbf{x}_i^c$ and all other
  proteins in $\mathbf{X}^c$;
  2. Identify the $k$ closest proteins;
  3. Impute the missing values of $x_i$ using the average of the corresponding
  $k$ closest.

Here, $\mathbf{x}_i^c$ and $\mathbf{X}_i^c$ denote the complete observed values
of $x_i$ and all other proteins.


```{r imputation}
assay(fts[["proteins"]], "log2_intensity") <- log2(assay(fts[["proteins"]]))
logx_imputed <- impute::impute.knn(
  data = assay(fts[["proteins"]], "log2_intensity"), k = 10L)
assay(fts[["proteins"]], "log2_imputed") <- logx_imputed$data
```

## Normalization

Normalization aims to remove systematic differences due to experimental 
instrumentation, ensuring that any observed heterogeneity or differential
expression between samples is driven by biological and not technical biases.
For label-free proteomics quantification data @Valikangas2018 systematically
reviewed normalization methods using spike-in data sets. The authors showed that
the faster cyclic loess normalization introduced by @Ballman2004 gives well
performing results for proteomics data.
We use the faster cyclic loess for normalization on the imputed log2 intensity.
This method is implemented in `normalizeCyclicLoess` function with the argument 
`method = "fast"` from the `limma` package [@Ritchie2015].


```{r normalization}
assay(fts[["proteins"]], "log2_normalized") <- limma::normalizeCyclicLoess(
  x = assay(fts[["proteins"]], "log2_imputed"), method = "fast")
```


# Comparison {#sec-comparison}

This section shows a graphical comparison between the data set processed by
@Margalit2022 and our proposed processing.

```{r import-margalit, include=FALSE}
se_margalit <- readRDS("../data/se_processed.rds")
rownames(se_margalit) <- rowData(se_margalit)$protein__id

# Organizing the data
pivotting_data <- function(se, i = "log2_imputed") {
  tb <- assay(se, i) |>
    t() |>
    dplyr::as_tibble() |>
    dplyr::mutate(replicate = se$replicate,
                  group = se$group,
                  variable = i) |>
    tidyr::pivot_longer(cols = -c(replicate, group, variable),
                        names_to = "proteins",
                        values_to = "value") |>
    dplyr::select(proteins, group, replicate, variable, value)
  tb
}

data_pivotted <- dplyr::bind_rows(
  pivotting_data(se = fts[["proteins"]], i = "intensity"),
  pivotting_data(se = fts[["proteins"]], i = "log2_imputed"),
  pivotting_data(se = fts[["proteins"]], i = "log2_normalized"),
  pivotting_data(se = se_margalit, i = "log_intensity")) |> 
  dplyr::mutate(variable = dplyr::case_when(
    variable == "log_intensity" ~ "Margalit et. al (2022)",
    variable == "intensity" ~ "Unprocessed intensity",
    variable == "log2_imputed" ~ "Log2 intensity imputed",
    variable == "log2_normalized" ~ "Log2 intensity normalized"),
    variable = forcats::fct_relevel(
      factor(variable), "Margalit et. al (2022)", "Unprocessed intensity",
      "Log2 intensity imputed"))
```

@fig-densities shows the protein abundance densities coloured by group and
faceted according to the transformation performed. The Unprocessed intensity
panel shows a very highly skewed distribution. The other panels show the
transformed intensity where we can see that the distribution is more symmetric.
When comparing the two below panels we can observe the effect of normalization.


```{r all-densities}
#| fig-width: 8
#| label: fig-densities
#| fig-cap: Proteins abundance densities by transformation and group.
ggplot(data_pivotted, aes(x = value, fill = group, col = group)) +
  facet_wrap(~variable, scales = "free") +
  geom_density(alpha = 0.3) +
  geom_rug(show.legend = FALSE) +
  labs(x = "Abundance", y = "Density", fill = "", col = "")
```

@fig-densities_comparison gives a comparison between the densities by the
processing methods. We clearly see that the distribution by @Margalit2022
are spread our more than ours, which are more symmetric.

```{r densities-comparison}
#| fig-width: 8
#| label: fig-densities_comparison
#| fig-cap: Comparison of proteins abundance densities between the processing approaches.
data_pivotted |> 
  dplyr::filter(variable %in% c("Margalit et. al (2022)",
                         "Log2 intensity normalized"),
                proteins %in% rownames(se_margalit)) |> 
  ggplot(aes(x = value, fill = variable, col = variable)) +
  facet_wrap(~group) +
  geom_density(alpha = 0.4) +
  geom_rug(show.legend = FALSE) +
  labs(x = "Abundance", y = "Density", fill = "", col = "") +
  scale_x_continuous(breaks = scales::pretty_breaks(6)) +
  scale_y_continuous(breaks = scales::pretty_breaks(6))
```

Another graphical inspection between the data processing steps is provided by the
PCA plots in @fig-pca. From the second panel we can see that
the samples from the same group are closer, while the data process used by
@Margalit2022 only grouped the samples from Cipofloxaxin.

```{r pca-comparison}
#| fig-height: 8
#| label: fig-pca
#| fig-cap: PCA performed on imputed $\log_2$ intensity before and after normalization.
pca_non_normalized <- scater::calculatePCA(
  x = fts[["proteins"]], exprs_values = "log2_imputed")
pca_normalized <- scater::calculatePCA(
  x = fts[["proteins"]], exprs_values = "log2_normalized")
pca_margalit <- scater::calculatePCA(
  x = se_margalit, exprs_values = "log_intensity")

tb <- dplyr::as_tibble(pca_non_normalized) |>
  dplyr::mutate(sample = rownames(pca_non_normalized),
                group = colData(fts[["proteins"]])$group,
                measure = "Log2 intensity imputed") |>
  dplyr::bind_rows(
    dplyr::as_tibble(pca_normalized) |>
      dplyr::mutate(group = colData(fts[["proteins"]])$group,
                    measure = "Log2 intensity normalized")) |>
  dplyr::bind_rows(
    dplyr::as_tibble(pca_margalit) |>
      dplyr::mutate(group = colData(se_margalit)$group,
                    measure = "Margalit et. al (2022)"))

ggplot(tb, aes(x = PC1, y = PC2, col = group)) +
  facet_wrap(~measure, ncol = 1) +
  geom_point(size = 4) +
  labs(x = "PCA 1", y = "PCA 2", col = "") +
  scale_x_continuous(breaks = scales::pretty_breaks(8)) +
  scale_y_continuous(breaks = scales::pretty_breaks(8))
```


```{r saving-fts, include=FALSE}
saveRDS(object = fts, file = "../data/fts_processed.rds")
```
