---
title: "Differential expression analysis"
author: "André F. B. Menezes, Beatrice Tropea, Cormac Monaghan"
date: "`r format(Sys.time(), '%B %d, %Y')`"
toc: true
number-sections: true
bibliography: references.bib
format:
  html:
    html-math-method: katex
    code-tools: true
    self-contained: true
    code-fold: true
    code-summary: "Show the code"
execute:
  warning: false
---

```{r pkg, include=FALSE}
library(QFeatures)
library(ggplot2)
library(ggrepel)
library(cowplot)
theme_set(
  theme_cowplot() +
    background_grid() +
    theme(legend.position = "top")
)
```


# Introduction {#sec-intro}

In this report we will perform a differential expression analysis in order to
find and ranking proteins in terms of evidence for differential expression (DE)
between the control group and the treatments
(Ampicillin, Cefotaxime, Impipenem, Ciprofloxacin).
We also perform deep investigation of the estimated $\log_2$ fold-change
between @Margalit2022 and the approached proposed.

The plan of this report is as follows. @sec-alternative_approach provides a
reviewed of moderated $t$-statistic proposed by @Smyth2004 and showed the
obtained results. @sec-differences compares the difference in terms of 
$\log_2$ fold-change and p-value between @Margalit2022 and the moderated
$t$-statistic. A further investigation is presented in @sec-top_3_de, where
the top 3 most DE proteins are examined.



```{r import-data}
fts <- readRDS(file = "../data/fts_processed.rds")
se_fiona <- readRDS("../data/se_processed.rds")
```

# Linear model {#sec-alternative_approach}

To test for the differential expression of proteins @Margalit2022 used the
pairwise Student-t test. It is well known that Student-t test has several
drawbacks in small sample size, for instance provides high false discovery
rates.
In order to overcome this limitation we will use the moderated $t$-statistic
introduced by @Smyth2004. This method was proposed in the context of microarray
experiment, though it can be easily adapted for label-free MS/MS data, since
the response variable, $\log_2$ normalized intensity usually can be well
approximated by a Normal distribution.


## The moderated $t$-statistic

The moderated $t$-statistic is derived from a hierarchical linear model,
defined by the existence of second order moments as follows

$$
\mathbb{E}(\mathbf{y}_{g}) = \mathbf{X}\,\boldsymbol{\alpha}_g
$$
and
$$
\mathrm{Var}(\mathbf{y}_{g}) = \mathbf{W}_g\,\sigma^2_g
$$

where $\mathbf{y}_{g} = (y_{g1}, \ldots, y_{gn})$ is the response variable 
vector (usually $\log_2$ normalized intensity values) for the $g$th protein,
$\mathbf{X}$ is a $p \times 1$ design matrix of full columns rank,
$\boldsymbol{\alpha}_g = (\alpha_{g1}, \ldots, \alpha_{gp})$ is the coefficient
vector,
$\mathbf{W}_g$ is a known non-negative definite weight matrix, and
$\sigma^2_g > 0$ is the common protein variance.

Usually the design matrix contains dummies variables indicating which group
the sample belongs. After the estimate $\boldsymbol{\alpha}_g$ the biological
interest is to test hypothesis about certain contrast of
$\boldsymbol{\alpha}_g$, without loss of generality we can defined the null
hypothesis as follows
$$
\mathcal{H}_0: \boldsymbol{\beta}_g = \mathbf{0}
$$
where $\boldsymbol{\beta} = \mathbf{c}^\top \boldsymbol{\alpha}_g$.

For example, the design matrix in this study is given by

```{r print-design}
# Creating the design matrix
g <- factor(colData(fts)$group)
g <- relevel(g, ref = "Control")
design_matrix <- model.matrix(~ g)
colnames(design_matrix) <- gsub("g", "", colnames(design_matrix))
design_matrix
```

Observe that the Control group is taken as reference level. Therefore, a matrix
contrast to test all pairwise comparison against the Control control is given
by
```{r}
mat_contrast <- matrix(c(-1, 1, 0, 0, 0,
                         -1, 0, 1, 0, 0,
                         -1, 0, 0, 1, 0,
                         -1, 0, 0, 0, 1),
                       nrow = 4, byrow = TRUE)
colnames(mat_contrast) <- c("Control", "Ampicillin", "Cefotaxime", "Impipenem",
                            "Ciprofloxacin")
rownames(mat_contrast) <- paste0("Contrast ", 1L:4L)
mat_contrast
```

For each protein $g = 1,\ldots, k$ the linear model is fitted to get the 
parameter estimators $\widehat{\mathbf{\alpha}}_g$,
$\widehat{s}_g$ of $\sigma_g^2$ and the covariance matrices

$$
\mathrm{Var}(\boldsymbol{\alpha}_{g}) = \mathbf{V}_g\,s_g^2,
$$
and
$$
\mathrm{Var}(\boldsymbol{\beta}_{g}) = \mathbf{c}\,\mathbf{V}_g\,\mathbf{c}^\top\,s_g^2
$$

An important remark is that the response variable is not necessarily normally
distributed and the parameters can be estimated with other method rather than
ordinary least squares. Let $\beta_{jg}$ a specific $j$ contrast of protein 
$g$, the distribution assumptions made by @Smyth2004 are in the estimators,
i.e.,

$$
\begin{align}
\widehat{\beta}_{gj}  \mid \beta_{gj}, \sigma^2_g &\sim N[\beta_{gj}, v_{gj}\,\sigma_g^2] \\
s_g^2 \mid \sigma_g^2 &\sim \dfrac{\sigma_g^2}{d_g}\,\chi^2_{d_g}
\end{align}
$$
where $v_{gj}$ is the $j$th diagonal element of
$\mathbf{c}\,\mathbf{V}_g\,\mathbf{c}^\top$, $d_g$ is the residual degree of
freedom for the linear model of protein $g$.

Other interesting aspect developed by @Smyth2004 is the possibility of borrow
information across proteins through use of prior information for the
parameters $\beta_{gj}$ and $\sigma_g^2$.

To describe how the variances are expected to vary across the proteins
@Smyth2004 assumes the following prior distribution for $\sigma_g^2$

$$
\dfrac{1}{\sigma_g^2} \sim \dfrac{1}{d_0\,s_0^2}\chi^2_{d_0}
$$
with a prior sample variance $s_0^2$ and $d_0$ degrees of freedom.

Important note that @Smyth2004 specify a conjugate prior for the sample
variance $s^2_g$ given $\sigma^2_g$, which can be summarized as:

$$
\begin{align}
(s^2_g \mid \sigma_g^2) &\sim \textrm{Gamma}\left[\dfrac{d_g}{2} , \dfrac{d_g}{2\,\sigma_g^2}\right] \\
\dfrac{1}{\sigma^2_g} &\sim \textrm{Gamma}\left[\dfrac{d_0}{2} , \dfrac{d_0\,s_0^2}{2}\right]
\end{align}
$$
where $X \sim \textrm{Gamma}(a, b)$ means gamma distribution with shape parameter
$a$, rate parameter $b$ and expected values $\mathbb{E}(X) = \dfrac{a}{b}$. 

Note that the $\mathbb{E}\left[\dfrac{1}{\sigma^2_g}\right] = \dfrac{1}{s_0^2}$.
Also, note that the shape parameter is known, then this conjugate prior leads
to the following posterior distribution

$$
\left(\dfrac{1}{\sigma_g^2}  \mid s^2_g \right) \sim \textrm{Gamma}\left[\dfrac{d_0 + d_g}{2} , \dfrac{d_0\,s_0^2 + d_g\,s_g^2}{2}\right]
$$
which has the following posterior mean:
$$
\mathbb{E}\left(\dfrac{1}{\sigma_g^2}  \mid s^2_g \right) = \dfrac{1}{\widetilde{s}^2_g}= \dfrac{d_0 + d_g}{d_0\,s_0^2 + d_g\,s_g^2}
$$
Hence, @Smyth2004 defined the moderated $t$-statistic for protein $g$ by

$$
\widetilde{t}_{gj} = \dfrac{\widehat{\beta}_{gj}}{\widetilde{s}_g\,\sqrt{v_{gi}}}
$$
where the posterior variance has replaced the usual sample variance of
ordinary $t$-statistic.

@Smyth2004 showed that under the null hypothesis the asymptotic distribution of 
$\widetilde{t}_{gj}$ is t-Student with $d_g + d_0$ degrees of freedom.
The additional degrees of freedom in $\widetilde{t}_{gj}$ reflects the extra
information borrow between the proteins.

From this point, the model assume that $d_0$ and $s_0^2$ are known. @Smyth2004
proposes an empirical Bayes approach to estimate these parameters from the
data. Particularly, $d_0$ and $s_0^2$ are estimated using the method of moments
and the distribution of $s_g^2$, which is shown to follows a a scaled
$F$-distribution.

The moderated t-statistic is available in `limma` package throughout the
`eBayes` function.

```{r de-analysis}
fit <- limma::lmFit(object = assay(fts[["proteins"]], "log2_normalized"),
                    design = design_matrix)
fit <- limma::eBayes(fit)
ampicillin <- limma::topTable(fit, coef = "Ampicillin", number = Inf,
                              sort.by = "none", confint = FALSE)
cefotaxime <- limma::topTable(fit, coef = "Cefotaxime", number = Inf,
                              sort.by = "none", confint = FALSE)
impipenem <- limma::topTable(fit, coef = "Impipenem", number = Inf,
                             sort.by = "none", confint = FALSE)
ciprofloxacin <- limma::topTable(fit, coef = "Ciprofloxacin", number = Inf,
                                 sort.by = "none", confint = FALSE)

# Appending
data_de_limma <- dplyr::bind_rows(
    dplyr::mutate(ampicillin, group = "Ampicillin",
                  protein = rownames(ampicillin)),
    dplyr::mutate(cefotaxime, group = "Cefotaxime",
                  protein = rownames(cefotaxime)),
    dplyr::mutate(impipenem, group = "Impipenem",
                  protein = rownames(impipenem)),
    dplyr::mutate(ciprofloxacin, group = "Ciprofloxacin",
                  protein = rownames(ciprofloxacin))) |> 
  dplyr::as_tibble() |> 
  dplyr::mutate(method = "limma") |> 
  dplyr::select(method, protein, group, logFC, P.Value, adj.P.Val) |> 
  dplyr::rename(log_fc = logFC, p_value = P.Value, fdr = adj.P.Val)

# Fiona's Student t-test --------------------------------------------------
expr_mat <- assay(se_fiona)
list_student_t <- lapply(seq_len(nrow(expr_mat)), function(i) {
  x_contrl <- expr_mat[i, colData(se_fiona)$group == "Control"]
  x_ampici <- expr_mat[i, colData(se_fiona)$group == "Ampicillin"]
  x_cefota <- expr_mat[i, colData(se_fiona)$group == "Cefotaxime"]
  x_impipe <- expr_mat[i, colData(se_fiona)$group == "Impipenem"]
  x_ciprof <- expr_mat[i, colData(se_fiona)$group == "Ciprofloxacin"]
  tibble::tibble(
    protein = rowData(se_fiona)$protein__id[i],
    group = c("Ampicillin", "Cefotaxime", "Impipenem", "Ciprofloxacin"),
    log_fc = c(mean(x_ampici - x_contrl), mean(x_cefota - x_contrl),
               mean(x_impipe - x_contrl), mean(x_ciprof - x_contrl)),
    p_value = c(t.test(x_contrl, x_ampici, var.equal = TRUE)$p.value,
                t.test(x_contrl, x_cefota, var.equal = TRUE)$p.value,
                t.test(x_contrl, x_impipe, var.equal = TRUE)$p.value,
                t.test(x_contrl, x_ciprof, var.equal = TRUE)$p.value)
  )
})

data_de_fiona <- do.call(rbind, list_student_t) |> 
  dplyr::mutate(method = "student_t") |> 
  dplyr::group_by(group) |> 
  dplyr::mutate(fdr = p.adjust(p_value, method = "fdr")) |> 
  dplyr::ungroup() |> 
  dplyr::select(method, protein, group, log_fc, p_value, fdr)


# Appending results -------------------------------------------------------
data_de <- dplyr::bind_rows(data_de_limma, data_de_fiona) |> 
  dplyr::mutate(method_label = ifelse(method == "student_t",
                                      "Margalit et al. (2022)",
                                      "Processed data + Limma"))

data_de_pivotted <- data_de |> 
  dplyr::group_by(group) |> 
  dplyr::mutate(rank_pvalue = rank(p_value)) |> 
  dplyr::ungroup() |> 
  dplyr::select(-c(method_label, fdr)) |> 
  tidyr::pivot_wider(names_from = method,
                     values_from = c(log_fc, p_value, rank_pvalue)) |> 
  dplyr::filter(!(is.na(log_fc_limma) | is.na(log_fc_student_t))) |> 
  dplyr::mutate(diff_log_fc = abs(log_fc_student_t) - abs(log_fc_limma),
                diff_p_value = p_value_student_t - p_value_limma,
                diff_rank = rank_pvalue_student_t - rank_pvalue_limma)
```


## Volcano plots

```{r volcano-plot-function, include=FALSE}
volcano_plot <- function(data, cutoff_logfc = 0.5,
                         cutoff_pvalue = 0.005, top_de = 10) {
  data_labelled <- data |> 
    dplyr::filter(abs(log_fc) >= cutoff_logfc, p_value <= cutoff_pvalue) |> 
    dplyr::arrange(-abs(log_fc)) |> 
    head(top_de)
  p <- ggplot(data, aes(x = log_fc, y = -log10(p_value))) +
    geom_point(size = 2, shape = 21, col = "black", fill = "grey69") +
    geom_point(data = data_labelled, shape = 21, size = 2, fill = "red",
               colour = "black") +
    geom_text_repel(data = data_labelled, aes(label = protein), size = 4,
                    color = "black") +
    geom_vline(xintercept = c(-cutoff_logfc, cutoff_logfc), linetype = "dashed",
               col = "black") +
    geom_hline(yintercept = -log10(cutoff_pvalue), linetype = "dashed",
               col = "black") +
    labs(x = expression("log"[2]*"FC"), y = expression("-log"[10]*"p-value")) +
    scale_x_continuous(breaks = scales::pretty_breaks(8)) +
    scale_y_continuous(breaks = scales::pretty_breaks(6))
    
  p
}
```

@fig-volcano-plot-ampicillin, @fig-volcano-plot-cefotaxime,
@fig-volcano-plot-impipenem and @fig-volcano-plot-ciprofloxacin show the
volcano plots obtained from moderated $t$-statistics for Control versus
Ampicillin, Cefotaxime, Impipenem, and Ciprofloxacin, respectively.
The cutoffs used for the p-value and the $\log_2$ fold-change were
0.005 and 0.5, respectively. We also colored in red and labelled the top 10
proteins ranked by the $\log_2$ fold-change.


```{r vp-ampicillin, fig.width=8}
#| label: fig-volcano-plot-ampicillin
#| fig-cap: Volcano plot for Control versus Ampicillin
volcano_plot(data = dplyr::filter(data_de_limma, group == "Ampicillin"))
```

```{r vp-cefotaxime, fig.width=8}
#| label: fig-volcano-plot-cefotaxime
#| fig-cap: Volcano plot for Control versus Cefotaxime
volcano_plot(data = dplyr::filter(data_de_limma, group == "Cefotaxime"))
```

```{r vp-impipenem, fig.width=8}
#| label: fig-volcano-plot-impipenem
#| fig-cap: Volcano plot for Control versus Impipenem
volcano_plot(data = dplyr::filter(data_de_limma, group == "Impipenem"))
```

```{r vp-ciprofloxacin, fig.width=8}
#| label: fig-volcano-plot-ciprofloxacin
#| fig-cap: Volcano plot for Control versus Ciprofloxacin
volcano_plot(data = dplyr::filter(data_de_limma, group == "Ciprofloxacin"))
```


# Comparing the differences {#sec-differences}

```{r tbl-pct-diff}
#| label: tbl-pct_diff
#| tbl-cap: Percentage of times that the difference in $\log_2$ fold-change and p-values between the methods are lesser than zero.

data_de_pivotted |>
  dplyr::group_by(group) |>
  dplyr::summarise(
    "Diff. log-FC" = paste0(round(100 * mean(diff_log_fc < 0), 3), "%"),
    "Diff. p-value" = paste0(round(100 * mean(diff_p_value < 0), 3), "%")) |> 
  kableExtra::kbl() |>
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r fig-diff-logfc}
#| label: fig-diff_logfc
#| fig-cap: Distributionof the difference in the estimated $\log_2$ fold change.

ggplot(data_de_pivotted, aes(x = diff_log_fc)) +
  facet_wrap(~group) +
  geom_density() +
  geom_rug() +
  geom_vline(xintercept = 0, col = "red") +
  labs(x = "Difference in the estimated log fold change", y = "Density") +
  scale_x_continuous(breaks = scales::pretty_breaks(6)) +
  scale_y_continuous(breaks = scales::pretty_breaks(4))
```

```{r fig-diff-pvalue}
#| label: fig-diff_pvalue
#| fig-cap: Distribution of the difference in the estimated p-value.

ggplot(data_de_pivotted, aes(x = diff_p_value)) +
  facet_wrap(~group) +
  geom_density() +
  geom_rug() +
  geom_vline(xintercept = 0, col = "red") +
  labs(x = "Difference in the p-value", y = "Density") +
  scale_x_continuous(breaks = scales::pretty_breaks(6)) +
  scale_y_continuous(breaks = scales::pretty_breaks(4))
```

```{r fig-total-de-proteins}
#| label: fig-total_de_proteins
#| fig-cap: Total of DE proteins according to the method and comparison.

data_de |> 
  dplyr::filter(p_value <= 0.005) |> 
  dplyr::group_by(method_label, group) |> 
  dplyr::count() |> 
  ggplot(aes(x = group, y = n, fill = method_label)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.6,
           col = "black") +
  geom_text(aes(label = n), vjust = -0.05,
            position = position_dodge(width = 0.9), size = 6) +
  labs(x = "", y = "", fill = "") +
  scale_y_continuous(limits = c(0, 55)) +
  theme(axis.line.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid = element_blank(),
        axis.text.y = element_blank())
```


# The top 3 most DE proteins {#sec-top_3_de}


## Comparing

```{r comparing-de, results="asis", fig.width=16}
groups <- unique(colData(fts)$group)
groups <- groups[which(groups != "Control")]
list_de <- list()
j <- 1L
for (g in groups) {
  cat("### ", g, "\n")

  most_de_fiona <- data_de_fiona |> 
    dplyr::filter(group == g, abs(log_fc) > 0.5) |> 
    dplyr::arrange(p_value) |> 
    head(3)
  
  most_de_limma <- data_de_limma |> 
    dplyr::filter(group == g, abs(log_fc) > 0.5) |> 
    dplyr::arrange(p_value) |> 
    head(3)

  data_curr <- data_de |> 
    dplyr::filter(group == g, protein %in% most_de_fiona$protein) |> 
    dplyr::mutate(most_de = "fiona") |> 
    dplyr::bind_rows(
      data_de |> 
        dplyr::filter(group == g, protein %in% most_de_limma$protein) |> 
        dplyr::mutate(most_de = "ours"))

  data_curr_f <- data_curr |> 
    dplyr::filter(most_de == "fiona")
  data_curr_o <- data_curr |> 
    dplyr::filter(most_de == "ours")
  
  # Limits
  y_lim_f <- c(min(data_curr_f$log_fc) - 0.25, max(data_curr_f$log_fc) + 0.25)
  y_lim_o <- c(min(data_curr_o$log_fc) - 0.25, max(data_curr_o$log_fc) + 0.25)
  
  # Plotting
  p_most_de_fiona <- ggplot(data_curr_f, aes(x = protein, y = log_fc,
                                             fill = method_label)) +
    geom_col(position = position_dodge(width = 0.9),
             alpha = 0.6, col = "black") +
    geom_hline(yintercept = 0, col = "black", size = 1.5) +
    geom_text(aes(label = paste0(round(log_fc, 2), " (",
                                 formatC(p_value, format = "e", digits = 1),
                                 ")")),
              position = position_dodge(width = 0.9), vjust = -0.2,
              size = 4) +
    labs(x = "Protein", y = "Log Fold Change", fill = "") +
    scale_y_continuous(breaks = scales::pretty_breaks(6), limits = y_lim_f) +
    ggtitle("3 most DE proteins by Margalit et al. (2022)")
  
  p_most_de_ours <- ggplot(data_curr_o, aes(x = protein, y = log_fc,
                                            fill = method_label)) +
    geom_col(position = position_dodge(width = 0.9),
             alpha = 0.6, col = "black") +
    geom_hline(yintercept = 0, col = "black", size = 1.5) +
    geom_text(aes(label = paste0(round(log_fc, 2), " (",
                                 formatC(p_value, format = "e", digits = 1),
                                 ")")),
              position = position_dodge(width = 0.9), vjust = -0.2,
              size = 4) +
    labs(x = "Protein", y = "Log Fold Change", fill = "") +
    scale_y_continuous(breaks = scales::pretty_breaks(6), limits = y_lim_o) +
    ggtitle("3 most DE proteins by Processed data + Limma")
  p_grided <- plot_grid(p_most_de_fiona, p_most_de_ours)

  print(p_grided)
    
  # Appending most DE proteins
  list_de[[j]] <- data.frame(
    group = g, fiona = most_de_fiona$protein, ours = most_de_limma$protein)
  j <- j + 1L

}
de_proteins <- do.call(rbind, list_de)
```



## The intensity distribution

```{r investigate, results="asis", fig.height=8}
col_data <- dplyr::as_tibble(colData(fts)) |> 
  dplyr::mutate(colname = rownames(colData(fts)))

# Function to pivot the raw intensity matrix into a long-format tibble
pivot_qfeatures <- function(fts, chosen_proteins) {
  list_tbs <- lapply(seq_len(length(fts)), function(j) {
    level <- names(fts)[j] 
    se_curr <- fts[[j]]
    chosen_rows <- rowData(se_curr)$protein %in% chosen_proteins
    
    i_assay <- if (level == "proteins") "intensity" else 1L
    row_data_curr <- dplyr::as_tibble(
      rowData(se_curr)[chosen_rows, "protein", drop = FALSE])
    
    pivotted <- assay(se_curr[chosen_rows, ], i_assay) |> 
      dplyr::as_tibble() |> 
      dplyr::bind_cols(row_data_curr) |> 
      tidyr::pivot_longer(cols = -c("protein"), names_to = "colname") |> 
      dplyr::mutate(level = level)
    pivotted
  })
  do.call(rbind, list_tbs) 
}

for (g in groups) {
  
  cat("### ", g, "\n")
  # Selecting the corresponding columns
  chosen_cols <- colData(fts)$group %in% c("Control", g)
  fts_curr <- fts[, chosen_cols]

  # Selecting the proteins
  most_de_limma <- de_proteins[de_proteins$group == g, "ours"]
  
  # Get the intensity values pivoted in a long format for all levels
  pivotted_limma <- pivot_qfeatures(fts = fts_curr,
                                    chosen_proteins = most_de_limma)
  
  # Join to get the columns data information
  pivotted_limma <- pivotted_limma |> 
    dplyr::left_join(col_data, by = "colname") |> 
    dplyr::select(level, protein, colname, group, replicate, sample_names,
                  value) |>
    dplyr::mutate(level = forcats::fct_relevel(
      factor(level), "psms", "peptides", "proteins"),
      group = forcats::fct_relevel(group, g)) |>
    dplyr::filter(level != "psms")
  
  # Plotting
  p_limma <- ggplot(pivotted_limma, aes(x = group, y = log2(value),
                                        colour = group)) +
    ggbeeswarm::geom_quasirandom(size = 2) +
    # geom_point(size = 2.5) +
    facet_wrap(protein ~ level, scales = "free", ncol = 2) +
    labs(x = "Group", y = "Log2 intensity", col = "") +
    ggtitle("Distribution of the 3 most DE by Processed data + Limma")
  print(p_limma)

}
```

```{r export, include=FALSE}
write.csv(x = data_de, file = "../data/de_proteins.csv", row.names = FALSE)
```



