---
title: "Feature selection analysis"
author: "André F. B. Menezes, Beatrice Tropea, and Cormac Monaghan"
date: "`r format(Sys.time(), '%B %d, %Y')`"
toc: true
number-sections: true
bibliography: references.bib
format:
  html:
    html-math-method: katex
    code-tools: true
    self-contained: true
    code-fold: true
    code-summary: "Show the code"
execute:
  warning: false
---

```{r setup}
library(ggplot2)
library(cowplot)
theme_set(
  theme_cowplot() +
    background_grid() +
    theme(legend.position = "top")
)
```


# Introduction {#sec-intro}

This document performs two methods of feature selection in order to reduce the
number of proteins by selecting the ones with high biological variability.
The first method is based on the decomposition of the variance into technical and
biological by nonparametrically modeling the variance intensity as a function of
mean intensity. The latter uses the Hedge's G effect size statistic.

All the analysis considers the processed $\log_2$ intensity matrix performed
in the processing steps described [here](processing/processing_steps.html).

```{r import-data, include=FALSE}
fts <- readRDS("../data/fts_processed.rds")
```


# Modeling mean-variance relationship {#sec-mean_variance}

This approach fits a locally-weighted regression (LOWESS) between the
mean-variance relationship of the $\log_2$ normalized intensity for all
proteins. We then subtract the variance of each protein by their corresponding
fitted values. The remaining values are called biological variability.
The `modelGeneVar` function from the `scran` package [@Aaron2016] is used.


```{r model-protein-var}
dec <- scran::modelGeneVar(x = fts[["proteins"]],
                           assay.type = "log2_normalized",
                           parametric = FALSE)

dec[order(dec$bio, decreasing = TRUE), ]
```

@fig-mean_var shows the mean-variance relationship with the fitted trend curve 
in blue. We observe two extreme points greater (lower) than 29 (22).

```{r visualizing}
#| fig-width: 8
#| label: fig-mean_var
#| fig-cap: Relationship between the mean-variance of the $\log_2$ intensity. Each point is a protein and the blue curve is the fitted model.
plot_mean_variance <- function(dec) {
  ggplot(dplyr::as_tibble(dec), aes(x = mean, y = total)) +
    geom_point(size = 2, alpha = 0.5) +
    geom_line(aes(y = tech), size = 1.5, col = "blue") +
    scale_x_continuous(breaks = scales::pretty_breaks(8)) +
    scale_y_continuous(breaks = scales::pretty_breaks(8)) +
    labs(x = "Mean of the log(intensity) normalized",
         y = "Variance of the log(intensity) normalized")
}
plot_mean_variance(dec = dec)
```

These two extreme proteins are listed below. Their inclusion could impact 
the fitted curve. Therefore, we will remove these cases from the analysis.

```{r remove-extreme-cases}
chosen <- ((dec$mean < 22) | (dec$mean > 29))
to_remove <- rownames(dec)[chosen]
to_remove
rowData(fts[["proteins"]][to_remove, ])
assay(fts[["proteins"]][to_remove, ], "intensity")
assay(fts[["proteins"]][to_remove, ], "log2_normalized")
```

After removing the extreme cases we re-fit the mean-variance trend curve.
The results are shown in @fig-mean_var_refitted. 
Note that 49.63% of the 

```{r refitting}
#| fig-width: 8
#| label: fig-mean_var_refitted
#| fig-cap: Relationship between the mean-variance of the $\log_2$ intensity after removing some proteins. Each point is a protein and the blue curve is the fitted model.
# Removing the two proteins with lower (greater) mean (m < 22 and m > 29) 
chosen <- !(rownames(fts[["proteins"]]) %in% to_remove)
fts[["proteins"]] <- fts[["proteins"]][chosen, ]

dec_2 <- scran::modelGeneVar(x = fts[["proteins"]],
                             assay.type = "log2_normalized",
                             parametric = FALSE)
dec_2[order(dec_2$bio, decreasing = TRUE), ]
table(dec_2$bio > 0)
plot_mean_variance(dec = dec_2)

# Concatenate the decomposition into SE object
if (all.equal(rownames(fts[["proteins"]]), rownames(dec_2)))
  rowData(fts[["proteins"]]) <- cbind(rowData(fts[["proteins"]]), dec_2)
```

```{r}
#| fig-width: 8
#| label: fig-ecdf-bio-var
#| fig-cap: Empirical cumulative distribution of the estimated biological variability.
ggplot(dplyr::as_tibble(dec_2), aes(x = bio)) +
  # geom_density(alpha = 0.4, fill = "grey68") +
  # geom_vline(xintercept = 0, col = "red") +
  stat_ecdf() +
  geom_point(data = data.frame(x = 0, y = mean(dec_2$bio > 0)),
             aes(x = x, y = y), col = "red", size = 4) +
  geom_rug() +
  labs(x = "Estimated biological variability", y = "Density") +
  scale_x_continuous(breaks = scales::pretty_breaks(8))
```


# Effect size statistic {#sec-effect_size}

For each comparison (Control vs Treatment) we can see which proteins are
up/down regulated in the control and remove them from the treatment because
it means that their expression is regulated by physiological gene expression.
Typically, Cohen's D is used for effect size calculation, however was noted to
be biased with regard to small sample sizes [@Bornstein2009]. To address this
we convert Cohen's D to Hedge's G by applying a correction factor [@Hedges1981]. 
The statistic is defined as followed, with the `hedges_g` function from the 
`effectsize` package [@Ben-Shachar2020] being used.

$$ 
G = \frac{(M_1 - M_2)}{\sqrt{\frac{{SD_1^2 + SD_2^2}}{2}}} \;\; (J) \;\;\; where \; J \approx (1 - \frac{3}{(4\; (n1 + n2) - 9})
$$

```{r effect-size}
ampicillin <- apply(
  assay(fts[["proteins"]], i = "log2_normalized"), 1,
  function(z)  effectsize::hedges_g(z[1:3], z[10:12])$Hedges_g)

cefotaxime <- apply(
  assay(fts[["proteins"]], i = "log2_normalized"), 1,
  function(x) effectsize::hedges_g(x[4:6], x[10:12])$Hedges_g)

ciprofloxacin <- apply(
  assay(fts[["proteins"]], i = "log2_normalized"), 1,
  function(x) effectsize::hedges_g(x[7:9], x[10:12])$Hedges_g)

impipenem <- apply(
  assay(fts[["proteins"]], i = "log2_normalized"), 1,
  function(x) effectsize::hedges_g(x[13:15], x[10:12])$Hedges_g)

# Organize the results in a data frame
effect_sizes <- data.frame(ampicillin, cefotaxime, ciprofloxacin, impipenem)
colnames(effect_sizes) <- paste0("hedges_g__", colnames(effect_sizes))
head(effect_sizes)

# Concatenating
if (all.equal(rownames(fts[["proteins"]]), rownames(effect_sizes)))
  rowData(fts[["proteins"]]) <- cbind(rowData(fts[["proteins"]]), effect_sizes)
```

@fig-densities_hedges_g shows the density of the absolute values of the Hedge's G
statistic according to the treatment. 

```{r densities-hedges-g}
#| fig-width: 8
#| label: fig-densities_hedges_g
#| fig-cap: Density of the absolute value of Hedges'G statistics compared with the control for each protein and according to the treatment.
pivotted_ef <- effect_sizes |>
  dplyr::mutate(protein = rownames(effect_sizes)) |> 
  tidyr::pivot_longer(cols = -protein, names_to = "hedges_g",
                      values_to = "value") |> 
  dplyr::mutate(variable = gsub("hedges_g__", "", hedges_g))
ggplot(pivotted_ef, aes(x = abs(value))) +
  facet_wrap(~variable) +
  geom_density(alpha = 0.4) +
  geom_rug() +
  labs(x = "Absolute value of Hedges'G", y = "Density") +
  scale_x_continuous(breaks = scales::pretty_breaks(8))
```

After computing the Hedge's G statistic we have to consider a threshold to
filter the most variability proteins. We will consider 0.5 as the threshold,
since this value means a moderate effect against the control.
@fig-densities_hedges_g displays the total number of proteins selected by group.

```{r checking}
#| fig-width: 8
#| label: fig-total_hedges_g
#| fig-cap: Total of selected proteins by group. The rule of thumb is the absolute value of Hedges greater than 0.5. 

total_filtered <- pivotted_ef |>
  dplyr::group_by(variable) |> 
  dplyr::summarise(n = sum(abs(value) > 0.5),
                   pct = mean(abs(value) > 0.5),
                   .groups = "drop")
ggplot(total_filtered, aes(x = reorder(variable, -n), y = n)) +
  geom_col() +
  geom_text(aes(label = paste0(n, " (", round(100 * pct, 2), "%)")),
            vjust = -0.4) +
  labs(x = "", y = "") +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank(),
        axis.ticks.y = element_blank())
```



# Comparison {#sec-comparison}

Finally, we plot in @fig-intersect-plot the intersection of proteins selected
between the two methods.

```{r intersect-plot}
#| fig-width: 8
#| label: fig-intersect-plot
#| fig-cap: Total of intersect proteins of the feature selection method by group.

chosen_var <- rownames(dec_2)[dec_2$bio > 0]
list_df <- list()
j <- 1L
for (g in unique(pivotted_ef$variable)) {
  chosen_hedges_g <- pivotted_ef[pivotted_ef$variable == g &
                                   abs(pivotted_ef$value) > 0.5, ]$protein
  list_df[[j]] <- data.frame(
    group = g, n = length(intersect(chosen_var, chosen_hedges_g)))
  j <- j + 1L
}
data_intersect <- do.call(rbind, list_df)
ggplot(data_intersect, aes(x = reorder(group, -n), y = n)) +
  geom_col() +
  geom_text(aes(label = n), vjust = -0.4) +
  labs(x = "", y = "") +
  theme(axis.text.y = element_blank(), axis.title.y = element_blank(),
        axis.ticks.y = element_blank(), panel.grid = element_blank())
```


```{r save-data}

```


